prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-03-19 inst/MyArticle/paper.Rmd
Edits in Method and Discussion
- code snippets
- text
- add links to packages
- add TODOs

---
title: prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - Witold E. Wolski^[Functional Genomics Center Zurich (FGCZ) - Swiss Federal Institute of Technology in Zurich \| University of Zurich]
  - Paolo Nanni$^*$
  - Jonas Grossmann$^*$
  - Maria d'Errico$^*$
  - Ralph Schlapbach$^*$
  - Christian Panse$^*$
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
  Protein mass spectrometry is widely used for quantitative proteomics
  studies and in particular for relative protein quantification.
  Nevertheless, there is still a need for a flexible and easy-to-use, 
  transparently supporting a variety of well principled statistical
  inference procedures data analysis application programmer interface
  in R. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates
  essential steps of the mass spectrometry-based label-free quantification
  data analysis workflow as: quality control, data normalization, protein
  aggregation, statistical modeling, hypothesis testing, and sample size
  estimation.
  We use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to visualize and model
  simple experimental designs with a single explanatory variable and complex
  experiments with multiple factors.
  Furthermore, we use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to benchmark
  data acquisition, data preprocessing or data modeling methods. Furthermore,
  we apply `r BiocStyle::Githubpkg("fgcz/prolfqua")` to develop highly
  customizable, visually appealing, and interactive data analysis reports
  in PDF or HTML format.
  The R package is available through https://github.com/fgcz/prolfqua.
---

@@

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

Several packages exist to model mass spectrometry LFQ experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features. For example, `r BiocStyle::Biocpkg("MSstats")` infers the model from the sample annotation and generates the model formula. At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level fold change and $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

@@

# Implementation & Methods (Experimental Section)

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in _MSStats_ format, only a few lines of code to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration are needed. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable (e.g., intensity) in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enabled us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported (e.g. auto-completion) and easy to use (OOP).

@@

```{r LFQData, echo=FALSE, fig.cap="(ref:LFQData)", out.width = '90%'}

knitr::include_graphics("LFQDataUML.png")

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

@@

## Benchmark dataset

Ionstar dataset, give Pride ids, describe MaxQuant settings and version. Give reference to Experiment hub and the _prolfquaData_ package.

@@

For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `lmerTest` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

If there are no observation in a group we assume that there are no observations because of the limit of detection (LOD). Furthermore, we assume that the variance of the protein is constant in all the conditions.
We impute the mean estimate of the group by using the average of $1\%$ smallest group mean intensities in the experiment. To estimate the variance in the condition, we use the pooled variance based on all conditions:

@@

To summarize peptide level models to protein models we did applied the method suggested by [@Suomi2017bEnhanced], that is to use the median scaled $p$-value of the peptide models and cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

To obtain the median $p$-value of the protein, we first rescaled the peptide $p$-values by taking the direction of the fold-change $\hat \beta$ into account, i.e.:

@@

Because we used the median as the i-th order statistic  $i = \frac{n}{2} + 0.5$. Therefore, $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$ are used to parameterize the CDF of the Beta distribution. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

# Example Analysis

The code snippets in this section demonstrate the basic usage of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package.
Peptide intensities are $\log_2$ transformed and robust z-score scaled but then rescaled to preserves the original range of the data using the method `robscale`. We show the distribution of the normalized intensities in Figure \@ref(fig:prepro) Panel A. Protein intensities are estimated using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. The boxplot (Panel D) shows the distribution of the protein intensity estimate in each condition. Finally, Panel C shows the distribution of the protein standard deviations for each condition.

@@

# Benchmarking

The peptide intensities from the MaxQuant _peptide.txt_ file were $log_2$ transformed, and subsequent robust z-score transformation, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e. proteins with a single peptide assignment. No further data filtering was used. We also did not use imputation. The protein level linear model was fitted to the protein intensities inferred from the peptide intensities using the Tukey's median polish. The mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model were fitted to the peptide intensities. The details of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` benchmark implementation are available online  [BenchmarkingIonstarData](https://wolski.github.io/prolfqua/articles/BenchmarkingIonstarData.html).

For benchmarking of MSstats we used the defaults specified in the [Benchmarking_MSstats](https://wolski.github.io/prolfqua/articles/)

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or $p$-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:FDRfdp)

The IonStar  [@shen2018ionstar] benchmark dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins the difference $\beta$ between two dilutions should be $\beta = 0$ while for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in small fold-changes $\beta = 1.2,1.25,1.3(3),1.5$.

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p$-value and moderated $p$-value. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate (TPR), false positive rate (FPR), or false discovery proportion (FDP) which are given by:

```{r confusionMatrix}
table <- data.frame( c("beta != 0", "beta == 0", "Total"), matrix(c("TP","FP","R","FN","TN","","P","N","m"), ncol = 3, byrow = T))
colnames(table) <- c("Prediction \\ Truth","E.coli", "H.sapiens", "Total")
knitr::kable(table, caption = "Confusion matrix, TP - true positive, FP - false positive, FN - false negative, P - all positive cases (all E. coli proteins), N - all negative cases (all H. sapiens proteins), m- all proteins.")

@@

```{r}
allBenchmarks <- readRDS("../../inst/Benchresults/allBenchmarks.RDS")
benchmark_msstats <- readRDS("../../inst/Benchresults/benchmark_msstats.RDS")
benchmark_prodA <- readRDS("../../inst/Benchresults/benchmark_medpolish_proDA.RDS")
msFragger <- readRDS(file = "../../inst/Benchresults/MSFragger_medpol_benchmark.RDS")

allBenchmarks$benchmark_mssstats <- benchmark_msstats
allBenchmarks$benchmark_msFragger <- msFragger
allBenchmarks$benchmark_proDA <- benchmark_prodA
allBenchmarks <- allBenchmarks[c("benchmark_imputation","benchmark_ProtModerated",  "benchmark_mixedModerated", "benchmark_ropeca","benchmark_merged","benchmark_mssstats","benchmark_proDA"   )]

@@

```{r benchmarkROC, fig.cap="(ref:benchmarkROC)", out.width = '90%', fig.width=12, fig.height=10}

ttt <- sapply(allBenchmarks, function(x){x$complete(TRUE)})

@@

A relevant parameter is the number of contracts a given method estimates from the data(see Figure \@ref(fig:FDRfdp) left Panel). It indicates how robust the procedure works in the presence of missing data. For each protein, we can estimate four contrast which for 4046 proteins, at least two peptides resulted in $16184$ possible contrast estimates. Because the set of contrasts slightly differs for each method, it makes comparing them complicated. For instance, we observe that MSstas estimates $16058$ while the mixed effect models estimate the fewest with $15940$. Hence, to conclude that one method shows a better performance, the score, e.g., pAUC, needs to be greater, but also the number of estimated contrasts needs to be equal or larger.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) right panel draws on the horizontal axis the FDR determined from the model, and on the vertical axis, the FDP obtained from the confusion matrix.  We see that most of the lines are below the diagonal, indicating that the FDR estimates obtained with `r BiocStyle::Githubpkg("fgcz/prolfqua")` are modestly conservative for this benchmark dataset provide. In the case of _MSstats_, we observe a high proportion of false discoveries for small $FDR$ values. In the case of `r BiocStyle::Biocpkg("PECA")`, the FDR estimates, obtained when Benjamini-Hochberg correcting the regulation probabilities, strongly overestimate the $FDP$.

@@

```{r FDRfdp, fig.cap = "(ref:FDRfdp)", out.width = '90%', fig.width=12, fig.height=10}

dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

@@

gridExtra::grid.arrange(p1_pAUC, p3, p4, layout_matrix=hlay)

@@

[Goes in the discussion or introduction, and give references for normalization e.g. VSN paper.] However, an essential parameter, not discussed further, is the normalization of the intensities within the samples to make them comparable. For this benchmark dataset, the fraction of regulated proteins (ECOLI) is relatively large. Therefore, to calibrate the measurement, we used human proteins only. In practice, it is unknown which proteins do and which don't change. Therefore, the assumption used when normalizing, scaling the data, and thus calibrating the measurement is that only a small subset of the proteins in the dataset is differentially regulated.

[Goes to discussion] The method used to infer protein intensities from more peptide intensities is an additional important factor. The original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. When using those, _proDA_ performs significantly worse than _proDA_ when using the protein estimates obtained from peptide intensity estimates using Tukey's median polish,  $pAUC_{10}(\textrm{t-statistics}) = 66%$ and $pAUC_{10}(\textrm{t-statistics}) = 73%$ respectively (see Articles on [@prolfquaGithubIO] for more details).

We assessed the performance of different modeling approaches implemented in prolfqua, MSstats and proDA, using a benchmark dataset whose ground truth was known. However, the dataset used has several limitations: first, the design is a parallel-group design, i.e. we could not benchmark how well interactions are modelled; secondly, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. The dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results.

@@

For the fixed effect linear model, the empirical Bayes variance shrinkage, as first suggested by [@Smyth2004linear], gives a consistent improvement compared with the unmoderated estimates (data not shown). However, there is no such effect for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for a large proportion of models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis. The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To adress the issue of missing observation we could either run proDA or the imputation of group averages method discussed here and afterwards apply `r BiocStyle::Biocpkg("PECA")`.

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values and $t$-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

>>>>

---
title: prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - Witold E. Wolski^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich]
  - Paolo Nanni$^*$
  - Jonas Grossmann$^*$
  - Maria d'Errico$^*$
  - Ralph Schlapbach$^*$
  - Christian Panse$^*$
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
  Protein mass spectrometry is widely used for quantitative proteomics
  studies and in particular for relative protein quantification.
  Nevertheless, there is still a need for a flexible and easy-to-use, 
  transparently supporting a variety of well principled statistical
  inference procedures data analysis application programmer interface
  in R. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates
  essential steps of the mass spectrometry-based label-free quantification
  data analysis workflow as: quality control, data normalization, protein
  aggregation, statistical modeling, hypothesis testing, and sample size
  estimation.
  We use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to visualize and model
  simple experimental designs with a single explanatory variable and complex
  experiments with multiple factors.
  Furthermore, we use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to benchmark
  data acquisition, data preprocessing or data modeling methods. Furthermore,
  we apply `r BiocStyle::Githubpkg("fgcz/prolfqua")` to develop highly
  customizable, visually appealing, and interactive data analysis reports
  in PDF or HTML format.
  The R package is available through https://github.com/fgcz/prolfqua.
---

@@

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested. Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

Several packages exist to model mass spectrometry LFQ experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features. For example, `r BiocStyle::Biocpkg("MSstats")` infers the model from the sample annotation and generates the model formula. At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a linear model formula and implements the empirical Bayes variance shrinkage method. The R-package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level fold change and $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage known from `r BiocStyle::Biocpkg("limma")`. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

@@

# Methods

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g. by auto-completion features, and easy to use (OOP).

@@

```{r LFQData, echo=FALSE, fig.cap="(ref:LFQData)", out.width = '90%'}
knitr::include_graphics("LFQDataUML.png")
```

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented (TODO(wew): cite the docu). This makes it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

## Dataset

TODO(wew): Ionstar dataset, give Pride ids, describe MaxQuant settings and version. Give reference to Experiment hub and the _prolfquaData_ package.

@@

For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

If there are no observation in a group, we assume that there are no observations because of the limit of detection (LOD). Furthermore, we assume that the variance of the protein is constant in all the conditions.
We impute the mean estimate of the group by using the average of $1\%$ smallest group mean intensities in the experiment. To estimate the variance in the condition, we use the pooled variance based on all conditions:

@@

To summarize peptide level models to protein models, we applied the method suggested by [@Suomi2017bEnhanced]  to use the median scaled $p$-value of the peptide models and cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

To obtain the median $p$-value of a protein, we first rescaled the peptide $p$-values by taking the direction of the fold-change $\hat \beta$ into account, i.e.:

@@

Because we used the median as the i-th order statistic $i = \frac{n}{2} + 0.5$. Therefore, $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$ are used to parameterize the (TODO(wew): what is CDF) CDF of the Beta distribution. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

## Usage

The code snippets in this section demonstrate the basic usage of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. As input data we use Ionstar dataset.
Peptide intensities are $\log_2$ transformed and robust z-score scaled but then rescaled to preserves the original range of the data using the method `robscale`. We show the distribution of the normalized intensities in Figure \@ref(fig:prepro) Panel A. Protein intensities are estimated using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. The boxplot (Panel D) shows the distribution of the protein intensity estimate in each condition. Finally, Panel C shows the distribution of the protein standard deviations for each condition.

@@

## Benchmarking

The peptide intensities from the _MaxQuant_ _peptide.txt_ file were $log_2$ transformed, and subsequent robust z-score transformation, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e. proteins with a single peptide assignment. No further data filtering was used. We also did not use imputation. The protein level linear model was fitted to the protein intensities inferred from the peptide intensities using the Tukey's median polish. The mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model were fitted to the peptide intensities. The details of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` benchmark implementation are available online  [BenchmarkingIonstarData](https://wolski.github.io/prolfqua/articles/BenchmarkingIonstarData.html).

For benchmarking of `r BiocStyle::Biocpkg("MSstats")` we used the defaults specified in the [Benchmarking_MSstats](https://wolski.github.io/prolfqua/articles/)

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g., the false discovery proportion FDP. We use those plots to study how well linear, mixed effect models or $p$-value moderation models quantitative mass spectrometric high throughput experiments and compare them with results produced by `r BiocStyle::Biocpkg("MSstats")` [@MSstats2014] and `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020] (see also Figure \@ref(fig:FDRfdp))

The IonStar [@shen2018ionstar] dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins, the difference $\beta$ between two dilutions should be $\beta = 0$. While for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in small fold-changes $\beta = (1.20, 1.25, 1.30, 1.50)$.

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p$-value and moderated $p$-value. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate ($TPR$), false positive rate ($FPR$), or false discovery proportion ($FDP$) which are given by:

```{r knitrConfusionMatrix}
table <- data.frame( c("beta != 0", "beta == 0", "Total"),
                     matrix(c("TP", "FP", "R", "FN", "TN", "", "P", "N", "m"),
                            ncol = 3, byrow = T))
colnames(table) <- c("Prediction \\ Truth","E.coli", "H.sapiens", "Total")

knitr::kable(table, caption = "Confusion matrix, TP - true positive, FP - false positive, FN - false negative, P - all positive cases (all E. coli proteins), N - all negative cases (all H. sapiens proteins), m- all proteins.")
```

with

@@

```{r readBenchmarkData}
allBenchmarks <- readRDS("../../inst/Benchresults/allBenchmarks.RDS")
benchmark_msstats <- readRDS("../../inst/Benchresults/benchmark_msstats.RDS")
benchmark_prodA <- readRDS("../../inst/Benchresults/benchmark_medpolish_proDA.RDS")
msFragger <- readRDS(file = "../../inst/Benchresults/MSFragger_medpol_benchmark.RDS")

allBenchmarks$benchmark_mssstats <- benchmark_msstats
allBenchmarks$benchmark_msFragger <- msFragger
allBenchmarks$benchmark_proDA <- benchmark_prodA
allBenchmarks <- allBenchmarks[c("benchmark_imputation","benchmark_ProtModerated",  "benchmark_mixedModerated", "benchmark_ropeca","benchmark_merged","benchmark_mssstats","benchmark_proDA"   )]
```

@@

```{r benchmarkROC, fig.cap="(ref:benchmarkROC)", out.width = '90%', fig.width=12, fig.height=10}
ttt <- sapply(allBenchmarks, function(x){x$complete(TRUE)})

@@

A relevant parameter is the number of contracts a given method estimates from the data(see Figure \@ref(fig:FDRfdp) left Panel). It indicates how robust the procedure works in the presence of missing data. For each protein, we can estimate four contrast which for 4046 proteins, at least two peptides resulted in $16184$ possible contrast estimates. Because the set of contrasts slightly differs for each method, it makes comparing them complicated. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ while the mixed effect models estimate the fewest with $15940$. Hence, to conclude that one method shows a better performance, the score, e.g., pAUC, needs to be greater, but also the number of estimated contrasts needs to be equal or larger.

@@

We also benchmark, if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) right panel draws on the horizontal axis the FDR determined from the model, and on the vertical axis, the FDP obtained from the confusion matrix. We see that most of the lines are below the diagonal, indicating that the FDR estimates obtained with `r BiocStyle::Githubpkg("fgcz/prolfqua")` are modestly conservative for this benchmark dataset provide. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of `r BiocStyle::Biocpkg("PECA")`, the FDR estimates, obtained when Benjamini-Hochberg correcting the regulation probabilities, strongly overestimate the $FDP$.

@@

```{r FDRfdp, fig.cap = "(ref:FDRfdp)", out.width = '90%', fig.width=12, fig.height=10}
dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

@@

gridExtra::grid.arrange(p1_pAUC, p3, p4, layout_matrix=hlay)
```

@@

[TODO(wew): Goes in the discussion or introduction, and give references for normalization e.g. VSN paper.] However, an essential parameter, not discussed further, is the normalization of the intensities within the samples to make them comparable. For this benchmark dataset, the fraction of regulated proteins (_E. coli_) is relatively large. Therefore, to calibrate the measurement, we used human proteins only. In practice, it is unknown which proteins do and which don't change. Therefore, the assumption used when normalizing, scaling the data, and thus calibrating the measurement is that only a small subset of the proteins in the dataset is differentially regulated.

[TODO(wew): Goes to discussion] The method used to infer protein intensities from more peptide intensities is an additional important factor. The original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. When using those, _proDA_ performs significantly worse than _proDA_ when using the protein estimates obtained from peptide intensity estimates using Tukey's median polish,  $pAUC_{10}(\textrm{t-statistics}) = 66%$ and $pAUC_{10}(\textrm{t-statistics}) = 73%$ respectively (see Articles on [@prolfquaGithubIO] for more details).

We assessed the performance of different modeling approaches implemented in prolfqua, `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`, using a benchmark dataset whose ground truth is known. However, the dataset used has several limitations: first, the design is a parallel-group design, i.e., we could not benchmark how well interactions are modelled; secondly, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. The dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results.

@@

For the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by [@Smyth2004linear], consistently improves compared with the unmoderated estimates (data not shown). However, there is no such effect for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for a large proportion of models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure TODO(wew)). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis. The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To address the issue of missing observation, we could either run `r BiocStyle::Biocpkg("proDA")` or the imputation of group averages method discussed here and afterwards apply `r BiocStyle::Biocpkg("PECA")`.

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values and $t$-statistics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to, e.g., repeated measurements. If you use fixed-effect linear models, apply variance moderation since this improves the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.
